The five axes of the Turtle: symmetry and asymmetry in NGC~6210

### 2020-12-07 ms-v1-expand.tex
Do latexdiff to get highlighted changes

\documentclass[useAMS, usenatbib]{mnras}
\usepackage{wjh-packages}
\usepackage{booktabs}
\hypersetup{colorlinks=True, linkcolor=blue!50!black, citecolor=black,
  urlcolor=blue!50!black}
\usepackage{etoolbox}
\robustify\bfseries
\robustify\itshape

@@

\begin{abstract}
  We carry out a comprehensive kinematic and morphological study
  of the asymmetrical planetary nebula:
  NGC~6210, known as the Turtle.
  The nebula's spectacularly chaotic appearance has led to proposals that it was shaped by mass transfer in a triple star system.
  We study the three-dimensional structure and kinematics
  of its shells, lobes, knots, and haloes
  by combining radial velocity mapping from multiple long-slit spectra
  with proper motion measurements from multi-epoch imaging.
  We find that the nebula has five distinct ejection axes.
  The first is the axis of the bipolar, wind-blown inner shell,
  while the second is the axis of the lop-sided, elliptical, fainter, but more massive intermediate shell.
  A further two axes are bipolar flows that form the point symmetric,
  high-ionization outer lobes, all with inclinations close to the plane of the sky.
  The final axis, which is inclined close to the line of sight,
  traces collimated outflows of low-ionization knots.
  We detect major changes in outflow directions during the planetary nebula phase, 
  starting at or before the initial ionization of the nebula 3500~years ago.
  Most notably,
  the majority of redshifted low-ionization knots have kinematic ages greater than 2000~years,
  whereas the majority of blueshifted knots have ages younger than 2000~years. 
  Such a sudden and permanent 180-degree flip in the ejection axis
  at a relatively late stage in the nebular evolution is a challenge to models of planetary nebula formation and shaping.
  % A comprehensive kinematic study has been carried out on the planetary nebula (PN) NGC 6210. Multiple long-slits, echelle spectra have been obtained over the face of this nebula mapping its full shell structure, the opposite pairs of extended, twisted arms and the bipolar collimated outflows and bullets. Public {\it HST} imagery has been used to identify kinematic elements with structural components. The long-slit spectroscopic information has been combined into channel maps that greatly facilitate visualizing the otherwise intricate expansion pattern of this  planetary nebula. The global morphology of NGC 6210 is  reminiscent of other planetary nebulae with extended X-ray emission, suggesting the presence of a central hot bubble produced by shocked stellar wind. In spite of the dramatic structure of this PN substantial expanding radial motions are only found in the material surrounding the central, inner shell. The relatively slow radial motions of the collimated outflows and the bullet-like knots indicate that they are  moving away from the core very close to the plane of the sky, nearly perpendicular to the line of sight. The direction and ages of the various components, derived from their expanding velocities indicate that NGC 6210 has been strongly influenced by the  action of a bipolar, rotating, episodic outflow.
\end{abstract}

@@

To the best of our knowledge,
the first use of the name Turtle with regard to NGC~6210
was in a 1998 NASA press release%
\footnote{\url{https://hubblesite.org/contents/news-releases/1998/news-1998-36.html}}
that described images of the nebula obtained with the
\textit{Hubble Space Telescope} (\textit{HST}).
Those and more recent \textit{HST} images%
\footnote{
  \url{https://www.spacetelescope.org/images/opo9836f/}
  and \url{https://www.spacetelescope.org/images/potw1026a/}.}
are shown in Figure~\ref{fig:hst},
with the most important morphological features labelled in the upper panels.
In most emission lines,
such as \Ha{} \Wav{6563} and \oiii{} \Wav{5007},
the brightest part of the nebula is the inner shell, which is elongated approximately NNW--SSE,
with a peanut-like shape.
The larger and fainter intermediate shell is elongated on a roughly perpendicular axis
and is not centered on the progenitor star, but is offset towards the WSW.
Outside these shells are the four lobes of the nebula, which form the ``flippers'' of the Turtle.
In low-ionization lines,
such as \nii{} \Wav{6583},
the shells emit only faintly and the lobes are not seen at all.
Instead, the brightest emission comes from a series of individual knots and knot complexes (seen as pink in Figure~\ref{fig:hst}d).
The N~knot is the farthest from the star and is superimposed on the N~lobe,
with no counterpart to the S.
The NW knot and SE knot form a symmetrical pair, although the SE knot is closer to the star.
The brightest of the more diffuse knot complexes are concentrated to the N and W of the star.

@@

\section{Longslit Spectroscopy}
\label{sec:observations}
 \begin{figure}
   \centering
% inside_import 
% before    
% ignored  
% args width=0.38\textwidth
% full_filename figs/turtle-slits
% after 
   \includegraphics[width=0.38\textwidth]{figs/turtle-slits}
   \caption{
     Spectrograph slit positions for vertical orientations (upper panel),
     horizontal orientations (middle panel),
     and diagonal orientations (lower panel).
     Background image in each case is an \textit{HST} \oiii{} image of the nebula.
     In the top panel, slit positions from the 2015 season are indicated as solid lines
     and positions from other seasons are shown as dashed lines (see Table~\ref{table:pa5} for details). 
   }
  \label{fig:slit-positions}
\end{figure}

\begin{table}
\centering
\caption{OAN MES-SPM high-resolution spectroscopy of NGC~6210}
\label{table:pa5}
\begin{tabular}{@{}l@{} rrrr l@{}} \toprule
  Epoch&   Exposure & Spectral  & Width & P.A.   & Slit Label\\
       &    (s) & Range  &  (\si{\um})    &    (\(^\circ\))  &  \\
  \addlinespace[1pt]
  {(1)} & {(2)} & {(3)} & {(4)} & {(5)} & {(6)} \\
  \midrule
1998/06/28 & 30 & \Ha  & 150 & 90 & o\\
1998/06/28 & 300 & \Ha  & 150 & 90 & n\\
1998/06/28 & 1200 & \Ha   & 150 & 90 & m,p,q \\
2003/06/05 & 1800 &  \Ha   & 70 &0 &     a,b,d*,e*,f,h,i,k\\ 
2003/10/16 &  1800 &   \Ha    & 70 & $-$21 & t\\
2003/10/16 &  1800 &   \Ha   & 70 & $-$68 & v\\
2003/10/17 &  1800 &   \Ha  & 70 & 77 & w\\
2004/06/13 & 1800 &   \oiii  & 70 & $-$9 & r  \\  
2004/06/14 & 1800 &  \sii &150 & $-$19 & s  \\
2004/06/13 & 1800 &   \oiii  & 70 & $-$19 & s  \\  
2004/06/13 & 1800 &   \oiii  & 70 & $-$56 & u  \\  
  2004/06/14 & 1800 &   \oiii  & 150 & 0 & a\(\prime\),l  \\
2011/05/21 & 1800 & \Ha  & 150 & 0 & g* \\
2011/05/21 & 600 & \oiii & 150 & 0  & g \\
2013/07/06 &  1800 & \Ha   & 150 & 0 & c,i,j \\
2015/08/18 &  1800 &   \Ha, \oiii  & 70 & 0 & c,d,e,f,g\\
2015/08/19 &  1800 &   \Ha, \oiii   & 70 & 0 & b,a*,i\\
2015/08/20 &  1800 &   \Ha, \oiii & 70 & 0 & h,j,k\\
2019/09/18 &  1800 &   \Ha    & 150 & 56 & x\\
  \bottomrule
  \addlinespace
  \multicolumn{6}{@{}p{\columnwidth}@{}}{
  \textsc{Columns:}
  (1)~Date of each observing run.
  (2)~Length of each spectroscopic exposure.
  (3)~Spectral range of each observation.  \Ha{} includes \nii{} \Wav{6548, 83} and \heii{} \Wav{6560}.  \sii{} includes both doublet lines \Wav{6716, 31}.
  (4)~Slit width in microns.
  (5)~Position angle of slit orientation in degrees.
  (6)~Labels assigned to each slit position, see Fig.~\ref{fig:slit-positions}.
  }
\end{tabular}
\end{table}

@@

\begin{figure}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-1d-line-profiles
% after 
  \includegraphics[width=\linewidth]{figs/turtle-1d-line-profiles}
  \caption{
    Thick colored lines show one-dimensional extracted velocity profiles
    of (top to bottom) \nii{} (red), \Ha{} (purple) and \oiii{} (light blue)
    for a  short section of slit~e
    (length \SI{2}{arcsec},
    centered \SI{3}{arcsec} north of the central star).
    For the \nii{} and \oiii{} lines, four-gaussian fits are shown
    (continuous thin black lines show the individual components,
    while thin dashed line shows the total profile).
    The bottom panel shows the residuals of these fits on an expanded scale.
    Short vertical colored lines mark the centroid velocity of each of the four components,
    with thicker lines corresponding to the high-ionization shell
    and thinner lines low-ionization knot complexes.
    The vertical dashed line shows the systemic velocity.
    No fits are performed to the \Ha{} line
    because the larger thermal broadening means that
    the four components cannot be reliably separated.
  }
  \label{fig:spec-1d}
\end{figure}

\begin{figure}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-halo-slit-x
% after 
  \includegraphics[width=\linewidth]{figs/turtle-halo-slit-x}
  \caption{
    Deep images and spectra of the outer halo of NGC~6210, showing knotty structure.
    (a)~Composite mosaic of \oiii{} images,
    each with an exposure time of 1800~s,
    giving a combined field of view of \(3' \times 3'\) centered on the nebula.
    The position of slit~x is shown, for which a deep \Ha{} spectrum was obtained.
    Note that different sections of the mosaic are shown with different brightness normalizations,
    leading to abrupt jumps at the boundaries.
    The images were taken with an occulting bar that blocks out the bright core of the nebula,
    allowing the faint halo to be seen.
    (b)~\Ha{} spectrum from slit~x, showing emission from two of the halo knots.
    The red dotted line shows the nebular nominal systemic velocity.
  }
  \label{fig:halo-knots}
\end{figure}

@@

All data was reduced using standard IRAF\footnote{IRAF was originally
  distributed by the National Optical Astronomy Observatory,
  \url{http://ast.noao.edu/data/software},
  and is currently available from \url{https://iraf-community.github.io}.}
routines.
Bias and cosmic-ray removal was followed by rectification and first-order wavelength 
calibration of the two-dimensional spectra
based on the comparison spectrum
of a Th/Ar lamp, yielding an accuracy in velocity units of \SI{\pm 1}{km.s^{-1}}.
Figure~\ref{fig:pv-array} shows example position-velocity arrays of (left-to-right)
\Ha{} \Wav{6563}, \oiii{} \Wav{5007}, and \nii{} \Wav{6583} for the 11 individual positions from the 2015 run.
The velocity scale is heliocentric 
and the scale along the slit is specified as arcsecond offsets from the central star.
Figure~\ref{fig:spec-1d} shows example one-dimensional spectra,
extracted from a short section of slit~e.
% We derive a heliocentric
% systemic velocity, \(\vsys = \SI{40}{km.s^{-1}}\) by using the slit position F (epoch 2003) that passes through the central star and is not saturated. 
An additional slit (x, shown in Figure~\ref{fig:halo-knots}) was placed on the northern outskirts of the nebula 
in the outer halo of NGC~6210 where filaments had been clearly detected in some of our deep images.\footnote{For this spectrum only, \(3 \times 3\) pixel on-chip binning was used.}
% The observation was carried out on 18 September 2019.  The spectrum was obtained in order 87 (\Ha{} and \nii lines) with the 150 micron slit oriented at a position angle of 56 degrees.  The detector was again the Marconi 2048x2048 with a 3$\times$3 binning.  The spectral dispersion was 0.087 \AA/pix and 0.525\arcsec/pixel.
Most of the spectra presented here are publicly available from the \mbox{San Pedro Mártir} Kinematic Catalogue of Galactic Planetary Nebulae\footnote{
  \url{http://kincatpn.astrosen.unam.mx}
}
\citep{Lopez:2012a}.

@@

\begin{figure*}
  \centering
  % \includegraphics[width=\linewidth]{figs/turtle-peanut-map-lores}
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-peanut-map
% after 
  \includegraphics[width=\linewidth]{figs/turtle-peanut-map}
  \caption{
    Velocity features in the high-ionization shells,
    which have been identified in the \oiii{} slits.
    (a)~Blue-shifted features. 
    (b)~Red-shifted features.
    Each are labelled with their line-of-sight velocity
    with respect to the nominal systemic velocity of \SI{-40}{km.s^{-1}}.
    Solid lines show features in the inner shell,
    dashed lines show features in the intermediate shell,
    and dot-dashed lines show miscellaneous features between the two shells.
    The line width is a qualitative indicator of the brightness of each feature.
    The background grayscale shows a high-pass filtered version of the HST \oiii{} image.
    The star symbol in this and subsequent figures indicates the position of the central star.
  }
  \label{fig:shell-velocity-components}
\end{figure*}

\begin{figure*}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-heii-shell-components
% after 
  \includegraphics[width=\linewidth]{figs/turtle-heii-shell-components}
  \caption{
    Velocity features in the highest-ionization gas,
    which have been identified in the \heii{} slits.
    (a)~Blue-shifted features. (b)~Red-shifted features.
    Each are labelled with their line-of-sight velocity
    with respect to the nominal systemic velocity of \SI{-40}{km.s^{-1}}.
    The line width is a qualitative indicator of the brightness of each feature,
    with dashed lines showing the very faintest features.
    The background grayscale image is the same as in Fig.~\ref{fig:shell-velocity-components}.
  }
  \label{fig:heii-shell-components}
\end{figure*}

\begin{figure*}
  \centering
  % \includegraphics[width=\linewidth]{figs/turtle-knot-complex-map-lores}
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-knot-complex-map
% after 
  \includegraphics[width=\linewidth]{figs/turtle-knot-complex-map}
  \caption{
    Velocity features in the low-ionization knot complexes,
    which have been identified in the \nii{} slits.
    (a)~Blue-shifted features.
    (b)~Red-shifted features.
    Each is labelled with their line-of-sight velocity
    with respect to the nominal systemic velocity of \SI{-40}{km.s^{-1}}.
    The line width is a qualitative indicator of the brightness of each feature.
  }
  \label{fig:knot-complex-map}
% \end{figure*}

% \begin{figure*}
%   \centering
  \bigskip
  \includegraphics[width=\linewidth]
  {figs/turtle-lobes-simplified-components}
  % {figs/turtle-lobes-simplified-components-lores}
  \caption{
    Velocity components in the outer lobes,
    which have been identified in the \oiii{} slits.
    (a)~Blue-shifted components identified in individual slits.
    (b)~Red-shifted components identified in individual slits.
  }
  \label{fig:outer-lobe-components}
\end{figure*}

\begin{figure}
  \centering
  % \includegraphics[width=\linewidth]{figs/turtle-halo-map-lores}
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-halo-map
% after 
  \includegraphics[width=\linewidth]{figs/turtle-halo-map}
  \caption{
    Velocity components in the halo, all measured from the \oiii{} spectra with slit positions as indicated.
    Components are divided into three classes:
    inner halo (pink ellipses), outer halo (yellow ellipses),
    and halo knots (green ellipses).
    The inner portion of the background image is
    the same high-pass-filtered \textit{HST} image
    shown in Figs.~\ref{fig:shell-velocity-components} and~\ref{fig:outer-lobe-components}.
    The outer portion of the background image is a deep \oiii{} exposure of the halo
    obtained with MES-SPM in direct imaging mode with an occulting bar (see Fig.~\ref{fig:halo-knots}).
  }
  \label{fig:halo-components}
\end{figure}

@@

We employ archival emission line images from the Wide-Field Planetary Camera~2 \citetext{WFPC2, \citealp{Holtzman:1995a}} of the \textit{Hubble Space Telescope} (\textit{HST}) to derive plane-of-sky motions in the nebula.
The images, which we downloaded from the Barbara~A. Mikulski Archive for Space Telescopes,%
\footnote{MAST, \url{https://archive.stsci.edu/}}
come from observations made between 1997 (Cycle 6) and 2008 (Cycle~16) via proposal IDs 6792, 7501, 8773, and 11122.
Original images had the source centered on the
WFPC2 camera's WFC3 chip (first epoch)
or PC chip (second epoch). 
Images were aligned to the celestial coordinate frame using Astrodrizzle%
\footnote{\url{https://drizzlepac.readthedocs.io}},
which also corrected for the optical distortions of the camera
and remapped all the images to a common uniform square pixel grid
at the WFC resolution of \SI{0.1}{arcsec.pix^{-1}}. 
Proper motions were then estimated from pairs of images
separated by 10 to 12 years,
using the Fourier Local Correlation Tracking (FLCT) method \citep{Welsch:2004a, Fisher:2008a}\footnote{
  We used version 1.07 of FLCT, obtained from \url{http://cgem.ssl.berkeley.edu/cgi-bin/cgem/FLCT/home},
  together with version 1.04 of the Python wrapper pyflct,
  obtained from \url{https://github.com/PyDL/pyflct}.}
with a kernel width of 10~pixels.
Results are shown in Figures~\ref{fig:proper-motions-oiii} and~\ref{fig:proper-motions-nii} for \oiii{} (filter F502N) and \nii{} (filter F658N), respectively.
The resultant per-pixel motions between the two epochs are found to be of order \SI{0.5}{pix} (\(\approx \SI{5}{mas.yr^{-1}}\))
and these raw results were then corrected by applying a global shift to force the motion of the central star to be zero.
The systematic error from the global alignment of the two epochs
and residual uncertainties in the distortion correction
is estimated to be \SI{1.5}{mas.yr^{-1}},
which is expected to dominate the proper motion uncertainties in the brighter parts of the nebula.
In fainter and more featureless regions of the nebula, the proper motions are increasingly affected by random noise, 
which we indicate by partially transparent arrows in the figures, 
such as can be seen in parts of the lobes in Figure~\ref{fig:proper-motions-oiii}.
To convert the angular motions into transverse velocities, we assume a distance of \SI{2}{kpc},
so that \SI{10}{mas.yr^{-1}} is equivalent to \SI{95}{km.s^{-1}}.

>>>>

\documentclass[useAMS, usenatbib]{mnras}
\pdfsuppresswarningpagegroup=1
%
\usepackage[spanish,es-minimal,english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{newtxtext}
\usepackage[stix2,smallerops]{newtxmath}
\usepackage{booktabs}
\hypersetup{colorlinks=True, linkcolor=blue!50!black, citecolor=black,
  urlcolor=blue!50!black}
\usepackage{etoolbox}
\robustify\bfseries
\robustify\itshape

@@

\begin{abstract}
  We carry out a comprehensive kinematic and morphological study
  of the asymmetrical planetary nebula:
  NGC~6210, known as the Turtle.
  The nebula's spectacularly chaotic appearance has led to proposals that it was shaped by mass transfer in a triple star system.
  We study the three-dimensional structure and kinematics
  of its shells, lobes, knots, and haloes
  by combining radial velocity mapping from multiple long-slit spectra
  with proper motion measurements from multi-epoch imaging.
  We find that the nebula has five distinct ejection axes.
  The first is the axis of the bipolar, wind-blown inner shell,
  while the second is the axis of the lop-sided, elliptical, fainter, but more massive intermediate shell.
  A further two axes are bipolar flows that form the point symmetric,
  high-ionization outer lobes, all with inclinations close to the plane of the sky.
  The final axis, which is inclined close to the line of sight,
  traces collimated outflows of low-ionization knots.
  We detect major changes in outflow directions over the last 6000~years,
  beginning in the pre-planetary nebula phase
  and continuing through the ionization of the nebula, which began 3500~years ago.
  Most notably,
  all redshifted low-ionization knots have kinematic ages greater than 2000~years,
  whereas all blueshifted knots have ages younger than 2000~years. 
  Such a sudden and permanent 180-degree flip in the ejection axis
  at a relatively late stage in the nebular evolution is a challenge to models of planetary nebula formation and shaping.
  % A comprehensive kinematic study has been carried out on the planetary nebula (PN) NGC 6210. Multiple long-slits, echelle spectra have been obtained over the face of this nebula mapping its full shell structure, the opposite pairs of extended, twisted arms and the bipolar collimated outflows and bullets. Public {\it HST} imagery has been used to identify kinematic elements with structural components. The long-slit spectroscopic information has been combined into channel maps that greatly facilitate visualizing the otherwise intricate expansion pattern of this  planetary nebula. The global morphology of NGC 6210 is  reminiscent of other planetary nebulae with extended X-ray emission, suggesting the presence of a central hot bubble produced by shocked stellar wind. In spite of the dramatic structure of this PN substantial expanding radial motions are only found in the material surrounding the central, inner shell. The relatively slow radial motions of the collimated outflows and the bullet-like knots indicate that they are  moving away from the core very close to the plane of the sky, nearly perpendicular to the line of sight. The direction and ages of the various components, derived from their expanding velocities indicate that NGC 6210 has been strongly influenced by the  action of a bipolar, rotating, episodic outflow.
\end{abstract}

@@

More recent images, obtained with the \textit{Hubble Space Telescope} (\textit{HST})%
\footnote{
  \url{https://www.spacetelescope.org/images/opo9836f/}
  and \url{https://www.spacetelescope.org/images/potw1026a/}.}
are shown in Figure~\ref{fig:hst},
with the most important morphological features labelled in the upper panels.
In most emission lines,
such as \Ha{} \Wav{6563} and \oiii{} \Wav{5007},
the brightest part of the nebula is the inner shell, which is elongated approximately NNW--SSE,
with a peanut-like shape.
The larger and fainter intermediate shell is elongated on a roughly perpendicular axis
and is not centered on the progenitor star, but is offset towards the WSW.
Outside these shells are the four lobes of the nebula, which form the ``flippers'' of the Turtle.
In low-ionization lines,
such as \nii{} \Wav{6583},
the shells emit only faintly and the lobes are not seen at all.
Instead, the brightest emission comes from a series of individual knots and knot complexes (seen as pink in Figure~\ref{fig:hst}d).
The N~knot is the farthest from the star and is superimposed on the N~lobe,
with no counterpart to the S.
The NW knot and SE knot form a symmetrical pair, although the SE knot is closer to the star.
The brightest of the more diffuse knot complexes are concentrated to the N and W of the star.

@@

\section{Longslit Spectroscopy}
\label{sec:observations}
 \begin{figure}
   \centering
% inside_import 
% before    
% ignored  
% args width=0.38\textwidth
% full_filename tere-figs/Figure2a
% after 
   \includegraphics[width=0.38\textwidth]{tere-figs/Figure2a}
   \caption{
     Spectrograph slit positions for vertical orientations (upper panel),
     horizontal orientations (middle panel),
     and diagonal orientations (lower panel).
     Background image in each case is an \textit{HST} \oiii{} image of the nebula.
     In the top panel, slit positions from the 2015 season are indicated as solid lines
     and positions from other seasons are shown as dashed lines (see Table~\ref{table:pa5} for details). 
   }
  \label{fig:slit-positions}
\end{figure}

\begin{table}
\centering
\caption{OAN MES-SPM high-resolution spectroscopy of NGC~6210}
\label{table:pa5}
\begin{tabular}{@{}l@{} rrrr l@{}} \toprule
  Epoch&   Exposure & Spectral  & Width & P.A.   & Slit Label\\
     &    (s) & Range  &  (\si{\um})    &    (\(^\circ\))  &  \\
  \midrule
  {(1)} & {(2)} & {(3)} & {(4)} & {(5)} & {(6)} \\
    \addlinespace
1998/06/28 & 30 & \Ha  & 150 & 90 & O\\
1998/06/28 & 300 & \Ha  & 150 & 90 & N\\
1998/06/28 & 1200 & \Ha   & 150 & 90 & M,P,Q \\
2003/06/05 & 1800 &  \Ha   & 70 &0 &     A,B,D*,E*,F,H,I,K\\ 
2003/10/16 &  1800 &   \Ha    & 70 & $-$21 & T\\
2003/10/16 &  1800 &   \Ha   & 70 & $-$68 & V\\
2003/10/17 &  1800 &   \Ha  & 70 & 77 & W\\
2004/06/13 & 1800 &   \oiii  & 70 & $-$9 & R  \\  
2004/06/14 & 1800 &  \sii &150 & $-$19 & S  \\
2004/06/13 & 1800 &   \oiii  & 70 & $-$19 & S  \\  
2004/06/13 & 1800 &   \oiii  & 70 & $-$56 & U  \\  
2004/06/14 & 1800 &   \oiii  & 150 & 0 & A',L  \\
2011/05/21 & 1800 & \Ha  & 150 & 0 & G* \\
2011/05/21 & 600 & \oiii & 150 & 0  & G \\
2013/07/06 &  1800 & \Ha   & 150 & 0 & C,I,J \\
2015/08/18 &  1800 &   \Ha, \oiii  & 70 & 0 & C,D,E,F,G\\
2015/08/19 &  1800 &   \Ha, \oiii   & 70 & 0 & B,A*,I\\
2015/08/20 &  1800 &   \Ha, \oiii & 70 & 0 & H,J,K\\
2019/09/18 &  1800 &   \Ha    & 150 & 56 & X\\
  \bottomrule
  \addlinespace
  \multicolumn{6}{@{}p{\columnwidth}@{}}{
  \textsc{Columns:}
  (1)~Date of each observing run.
  (2)~Length of each spectroscopic exposure.
  (3)~Spectral range of each observation.  \Ha{} includes \nii{} \Wav{6548, 83} and \heii{} \Wav{6560}.  \sii{} includes both doublet lines \Wav{6716, 31}.
  (4)~Slit width in microns.
  (5)~Position angle of slit orientation in degrees.
  (6)~Labels assigned to each slit position, see Fig.~\ref{fig:slit-positions}.
  }
\end{tabular}
\end{table}

@@

\begin{figure}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-halo-slit-x
% after 
  \includegraphics[width=\linewidth]{figs/turtle-halo-slit-x}
  \caption{
    Deep images and spectra of the outer halo of NGC~6210, showing knotty structure.
    (a)~Composite mosaic of \oiii{} images,
    each with an exposure time of 1800~s,
    giving a combined field of view of \(3' \times 3'\) centered on the nebula.
    The position of slit position X is shown, for which a deep \Ha{} spectrum was obtained.
    Note that different sections of the mosaic are shown with different brightness normalizations,
    leading to abrupt jumps at the boundaries.
    The images were taken with an occulting bar that blocks out the bright core of the nebula,
    allowing the faint halo to be seen.
    (b)~\Ha{} spectrum from Slit~X, showing emission from two of the halo knots.
    The red dotted line shows the nebular nominal systemic velocity.
  }
  \label{fig:halo-knots}
\end{figure}

@@

All data was reduced using standard IRAF\footnote{IRAF was originally
  distributed by the National Optical Astronomy Observatory,
  \url{http://ast.noao.edu/data/software},
  and is currently available from \url{https://iraf-community.github.io}.}
routines.
Bias and cosmic-ray removal was followed by rectification and first-order wavelength 
calibration of the two-dimensional spectra
based on the comparison spectrum
of a Th/Ar lamp, yielding an accuracy in velocity units of \SI{\pm 1}{km.s^{-1}}.
Figure~\ref{fig:pv-array} shows example position-velocity arrays of (left-to-right)
\Ha{} \Wav{6563}, \oiii{} \Wav{5007}, and \nii{} \Wav{6583} for the 11 individual positions from the 2015 run.
The velocity scale is heliocentric 
and the scale along the slit is specified as arcsecond offsets from the central star.
% We derive a heliocentric
% systemic velocity, \(\vsys = \SI{40}{km.s^{-1}}\) by using the slit position F (epoch 2003) that passes through the central star and is not saturated. 
An additional slit (X, shown in Figure~\ref{fig:halo-knots}) was placed on the northern outskirts of the nebula 
in the outer halo of NGC~6210 where filaments had been clearly detected in some of our deep images.\footnote{For this spectrum only, \(3 \times 3\) pixel on-chip binning was used.}
% The observation was carried out on 18 September 2019.  The spectrum was obtained in order 87 (\Ha{} and \nii lines) with the 150 micron slit oriented at a position angle of 56 degrees.  The detector was again the Marconi 2048x2048 with a 3$\times$3 binning.  The spectral dispersion was 0.087 \AA/pix and 0.525\arcsec/pixel.
Most of the spectra presented here are publicly available from the \mbox{San Pedro Mártir} Kinematic Catalogue of Galactic Planetary Nebulae\footnote{
  \url{http://kincatpn.astrosen.unam.mx}
}
\citep{Lopez:2012a}.

@@

\begin{figure*}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-peanut-map-lores
% after 
  \includegraphics[width=\linewidth]{figs/turtle-peanut-map-lores}
  \caption{
    Velocity features in the high-ionization shells,
    which have been identified in the \oiii{} slits.
    Left panel shows blue-shifted features,
    while right panel shows red-shifted features,
    each labelled with their line-of-sight velocity
    with respect to the nominal systemic velocity of \SI{-40}{km.s^{-1}}.
    Solid lines show features in the inner shell,
    dashed lines show features in the intermediate shell,
    and dot-dashed lines show miscellaneous features between the two shells.
    The line width is a qualitative indicator of the brightness of each feature.
    The background grayscale shows a high-pass filtered version of the HST \oiii{} image.
  }
  \label{fig:shell-velocity-components}
\end{figure*}

\begin{figure*}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-heii-shell-components
% after 
  \includegraphics[width=\linewidth]{figs/turtle-heii-shell-components}
  \caption{
    Velocity features in the highest-ionization gas,
    which have been identified in the \heii{} slits.
    Left panel shows blue-shifted features,
    while right panel shows red-shifted features,
    each labelled with their line-of-sight velocity
    with respect to the nominal systemic velocity of \SI{-40}{km.s^{-1}}.
    The line width is a qualitative indicator of the brightness of each feature,
    with dashed lines showing the very faintest features.
    The background grayscale image is the same as in Fig.~\ref{fig:shell-velocity-components}.
  }
  \label{fig:heii-shell-components}
\end{figure*}

\begin{figure*}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-knot-complex-map-lores
% after 
  \includegraphics[width=\linewidth]{figs/turtle-knot-complex-map-lores}
  \caption{
    Velocity features in the low-ionization knot complexes,
    which have been identified in the \nii{} slits.
    Left panel shows blue-shifted features,
    while right panel shows red-shifted features,
    each labelled with their line-of-sight velocity
    with respect to the nominal systemic velocity of \SI{-40}{km.s^{-1}}.
    The line width is a qualitative indicator of the brightness of each feature.
  }
  \label{fig:knot-complex-map}
% \end{figure*}

% \begin{figure*}
%   \centering
  \bigskip
  \includegraphics[width=\linewidth]
  {figs/turtle-lobes-simplified-components-lores}
  \caption{
    Velocity components in the outer lobes,
    which have been identified in the \oiii{} slits.
    (a)~Blue-shifted components identified in individual slits.
    (b)~Red-shifted components identified in individual slits.
  }
  \label{fig:outer-lobe-components}
\end{figure*}

\begin{figure}
  \centering
% inside_import 
% before   
% ignored  
% args width=\linewidth
% full_filename figs/turtle-halo-map-lores
% after 
  \includegraphics[width=\linewidth]{figs/turtle-halo-map-lores}
  \caption{
    Velocity components in the halo, all measured from the \oiii{} spectra with slit positions as indicated.
    Components are divided into three classes:
    inner halo (pink ellipses), outer halo (yellow ellipses),
    and halo knots (green ellipses).
    The inner portion of the background image is
    the same high-pass-filtered \textit{HST} image
    shown in Figs.~\ref{fig:shell-velocity-components} and~\ref{fig:outer-lobe-components}.
    The outer portion of the background image is a deep \oiii{} exposure of the halo
    obtained with MES-SPM in direct imaging mode with an occulting bar (see Fig.~\ref{fig:halo-knots}).
  }
  \label{fig:halo-components}
\end{figure}

@@

We employ archival emission line images from the Wide-Field Planetary Camera~2 \citetext{WFPC2, \citealp{Holtzman:1995a}} of the \textit{Hubble Space Telescope} (\textit{HST}) to derive plane-of-sky motions in the nebula.
The images, which we downloaded from the Barbara~A. Mikulski Archive for Space Telescopes,%
\footnote{MAST, \url{https://archive.stsci.edu/}}
come from observations made between 1997 (Cycle 6) and 2008 (Cycle~16) via proposal IDs 6792, 7501, 8773, and 11122.
Images were aligned with Astrodrizzle%
\footnote{\url{https://drizzlepac.readthedocs.io}}
and proper motions were estimated from pairs of images seperated by 10 to 12 years,
% The data used in this study to measure the proper motions of the turttle nebula consist of images which were retrieved from the HST archive. The \nii{} observations were made in two separate observing runs: on 1998 June 30, as part of program 7501, with Hajian, Arsen R., as PI. During this run, six images were obtained with the WFPC2, with an exposure time of 100 s (three images) y 40 s (three images) using the F658N filter. The second-epoch data were obtained on 2008 January 18, from the science program 11122, with Bruce Balick, as PI, who employed the same configuration as in program 7501 to acquire three images with the F658N filter, using 400 s exposure times for all of them. For all of the images, we used the drizzled images (processed with Astrodrizzle). 
using the FLCT method \citep{Welsch:2004a, Fisher:2008a}.\footnote{
  We used version 1.07 of FLCT, obtained from \url{http://cgem.ssl.berkeley.edu/cgi-bin/cgem/FLCT/home},
  together with version 1.04 of the Python wrapper pyflct,
  obtained from \url{https://github.com/PyDL/pyflct}.}
Results are shown in Figures~\ref{fig:proper-motions-oiii} and~\ref{fig:proper-motions-nii} for \oiii{} (filter F502N) and \nii{} (filter F658N), respectively.
In both cases, the images were remapped to a uniform square pixel grid at the WFC resolution of \SI{0.1}{arcsec.pix^{-1}} before applying the algorithm.
The resultant per-pixel motions between the two epochs are found to be of order \SI{0.5}{pix} (\(\approx \SI{5}{mas.yr^{-1}}\))
and these raw results were then corrected by applying a global shift to force the motion of the central star to be zero.
The systematic error from the global alignment of the two epochs is estimated to be \SI{1.5}{mas.yr^{-1}},
which is expected to dominate the proper motion uncertainties in the brighter parts of the nebula.
In fainter and more featureless regions of the nebula, the proper motions are increasingly affected by random noise, 
which we indicate by partially transparent arrows in the figures, 
such as can be seen in parts of the lobes in Figure~\ref{fig:proper-motions-oiii}.
To convert the angular motions into transverse velocities, we assume a distance of \SI{2}{kpc},
so that \SI{10}{mas.yr^{-1}} is equivalent to \SI{95}{km.s^{-1}}.
